Active Learning for SAT Solver Benchmarking

### 2022-10-11 paper/paper.tex
Harmonize section labels
Use two-level names consistently. No visible changes in compiled PDF.

In this section, we present the details of our approach.
Section~\ref{sec:main-framework} describes the general framework.
Section~\ref{sec:main-model} specifies the details of the underlying prediction model and describes how we may derive a solver ranking from that model.
We discuss possible criteria for sampling instances in Section~\ref{sec:main-sampling}.
Section~\ref{sec:main-stopping} concludes with possible stopping conditions.

\subsection{Incremental Benchmarking Framework}
\label{sec:main-framework}

@@

  \While(\tcp*[f]{cf. Section~\ref{sec:main-stopping}}){$\operatorname{not} \, \operatorname{stop}\!\left(\mathcal{M}\right)$}{
    $s \leftarrow \operatorname{selectNextInstance}\!\left(\mathcal{M}\right)$ \tcp*{cf. Section~\ref{sec:main-sampling}}

@@

    $\operatorname{updateModel}\!\left(\mathcal{M},\, s,\, t\right)$ \tcp*{cf. Section~\ref{sec:main-model}}
  }

  \Return $\operatorname{predictScore}\!\left(\mathcal{M}\right)$ \tcp*{cf. Section~\ref{sec:main-model}}

@@

\subsection{Solver Model}
\label{sec:main-model}

@@

\subsection{Instance Sampling}
\label{sec:main-sampling}

@@

\subsection{Stopping Criteria}
\label{sec:main-stopping}

@@

\subsection{Optimization Goal}
\label{sec:goal}

As already stated in the introductory section, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
We create a prediction model $\mathcal{M}$ as described in Section~\ref{sec:main-model} that provides us with a scoring function $s_k$.
% MI: Let this be the evaluation result:
%Also, it outputs the fraction of runtime that is needed to arrive at its conclusion.
First and foremost, our goal is to provide the engineer of new SAT solvers with an accurate ranking.
We define the \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) by the fraction of pairs $\left(\hat{a}, a\right)$ for all $a \in \mathcal{A}$ that are decided correctly by the given ranking (cf.~Algorithm~\ref{alg:eval}).
For now, we exclude the equality of solvers since data shows that the PAR-2 scores of all our solvers are different.
So, possible ranking decisions may be, for example, solver $a$ is better than $b$ or $b$ is better than $a$.
The ranking accuracy is affected by whether this decision is correctly made.
%Matricon et. al.~\cite{MatriconAFSH21} also evaluate their approach by the fraction of solver pairs that is correctly decided.
Second, we also have to optimize for runtime.
The \emph{fraction of runtime} that the algorithm needs to arrive at its conclusion is denoted by $O_{\operatorname{rt}} \in \left[0, 1\right]$ (lower is better).
This metric puts the runtime summed over the sampled instances in relation to the runtime summed over all instances in the dataset (cf. Algorithm~\ref{alg:eval}).
Overall, we want to find an approach that maximizes
%
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
%
whereby $\delta \in \left[0, 1\right]$ allows for linear weighting between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
Plotting the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on a $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.

\subsection{Evaluation Framework}
\label{sec:evalframe}

@@

\subsection{Hyper-parameters}
\label{sec:hyper}

Given Algorithm~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}.
Also, there are different choices for the runtime-label prediction model.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results (cf. Section~\ref{sec:evalprel}) and do the main end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.

@@

Regarding \emph{ranking} (cf. Section~\ref{sec:main-model}), we experiment with the following approaches, including our used hyper-parameter values:

\begin{itemize}
  \item Observed PAR-2 ranking of already sampled instances
  \item Predicted ranking induced by runtime-label predictions
  \begin{itemize}
    \item
    History size: consider the latest 1, 10, 20, 30, or 40 predictions within a voting approach for stability.
    The latest $x$ predictions vote on the winner.
    %JB: confuses me a bit as it sounds like we rank first and then vote on a winner; isn't it the other way around?
    %TF: Removed confusing sentence
    \item
    Fallback threshold: if the difference of scores between two solvers drops below \SI{0.01}, \SI{0.05}, or \SI{0.1}, use the partially observed PAR-2 ranking as a tie-breaker.
    %JB: as clarification: one these two solvers has to be the new solver, right? if yes, would write it explicitly
    %TF: Not necessarily
    As discussed in Section~\ref{sec:main-model}, ranking decisions are all correct if the difference in scores between two solvers exceeds a certain value.
  \end{itemize}
\end{itemize}

@@

For \emph{selection} (cf. Section~\ref{sec:main-sampling}), we experiment with the following methods, including our used hyper-parameter values:

@@

For \emph{stopping} decisions (cf. Section~\ref{sec:main-stopping}), we experiment with the following criteria, including our used hyper-parameter values:

@@

As runtime-label prediction model, we only make use of one fixed model since an exhaustive grid search would be infeasible (cf.~Section~\ref{sec:evalprel}).
Our model of choice is an ensemble, stacking a quadratic-discriminant analysis\footnote{With a regularization parameter of zero and a singular value threshold of zero} onto a random forest model\footnote{With \emph{entropy} splitting criterion and \emph{balanced} class weights} using a decision tree\footnote{With \emph{Gini} impurity splitting criterion; choosing the best split for each branching decision and maximum tree depth of 5 levels} to weight its members.
If not stated explicitly, all other model parameters default to the presets of \emph{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/index.html}; Version 1.0.2}~\cite{scikit-learn}.

@@

\subsection{Runtime Prediction}
\label{sec:evalprel}

@@

For runtime discretization, we use hierarchical clustering with $k = 3$ and a log-single-link criterion for all further experiments.
Each non-time-out runtime starts in a separate interval.
We then gradually merge intervals whose single-link logarithmic distance is the smallest until the desired number of partitions is reached.
As already mentioned in Section~\ref{sec:main-model}, this approach has good properties regarding the predictiveness of the PAR-2 score ranking and its discriminatory power.
Also, experiments showed that our stacking ensemble performed particularly well for this kind of discretization method.
Other clustering approaches that we have tried include hierarchical clustering with mean-, median- and complete-link criteria, as well as $k$-means and spectral clustering.
%Predicting runtime labels in a similar experiment as above produced an MCC score of on average \SI{88.82}{\%} ($\sigma = \SI{4.75}{\%}$; in comparison to \SI{95.27}{\%} for predicting time-outs).
%JB: "similar experiment as above" - reference unclear; what do you want to say?
%TF: Removed the sentence

@@

Our end-to-end experiments follow the evaluation framework introduced in Section~\ref{sec:evalframe}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches from Section~\ref{sec:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the smaller hyper-parameter-tuning dataset.
Evaluating a particular configuration with Algorithm~\ref{alg:eval} returns a point $\left(O_{\operatorname{rt}},\, O_{\operatorname{acc}}\right)$.
The plotted lines represent the best performing configurations (convex hull) per top-level hyper-parameter choice, i.e., per ranking approach (Fig.~\ref{fig:annitraincolorranking}), selection approach (Fig.~\ref{fig:annitraincolorselection}), and stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).
%Colors indicate the different hyper-parameter instantiations.
%To be more specific, they represent the Pareto front of all configurations with the respective instantiations.

@@

  \caption{
  	$O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams comparing different hyper-parameter instantiations of our active-learning framework.
  	The x-axis shows the ratio of total solver runtime on the sampled instances relative to all instances.
  	The y-axis shows the ranking accuracy (cf. Section~\ref{sec:goal}).
  	%JB: we should use "ranking accuracy" in plots, since we explicitly defined this metric earlier
  	Each line entails the front of Pareto-optimal configurations for the respective hyper-parameter instantiation.
  }
  \label{fig:e2eallsolvers}
\end{figure}

>>>>

In this section, we present the details of our approach.
Section~\ref{sec:main:framework} describes the general framework.
Section~\ref{sec:main:model} specifies the details of the underlying prediction model and describes how we may derive a solver ranking from that model.
We discuss possible criteria for sampling instances in Section~\ref{sec:main:sampling}.
Section~\ref{sec:main:stopping} concludes with possible stopping conditions.

\subsection{Incremental Benchmarking Framework}
\label{sec:main:framework}

@@

  \While(\tcp*[f]{cf. Section~\ref{sec:main:stopping}}){$\operatorname{not} \, \operatorname{stop}\!\left(\mathcal{M}\right)$}{
    $s \leftarrow \operatorname{selectNextInstance}\!\left(\mathcal{M}\right)$ \tcp*{cf. Section~\ref{sec:main:sampling}}

@@

    $\operatorname{updateModel}\!\left(\mathcal{M},\, s,\, t\right)$ \tcp*{cf. Section~\ref{sec:main:model}}
  }

  \Return $\operatorname{predictScore}\!\left(\mathcal{M}\right)$ \tcp*{cf. Section~\ref{sec:main:model}}

@@

\subsection{Solver Model}
\label{sec:main:model}

@@

\subsection{Instance Sampling}
\label{sec:main:sampling}

@@

\subsection{Stopping Criteria}
\label{sec:main:stopping}

@@

\subsection{Optimization Goal}
\label{sec:exdesign:goal}

As already stated in the introductory section, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
We create a prediction model $\mathcal{M}$ as described in Section~\ref{sec:main:model} that provides us with a scoring function $s_k$.
% MI: Let this be the evaluation result:
%Also, it outputs the fraction of runtime that is needed to arrive at its conclusion.
First and foremost, our goal is to provide the engineer of new SAT solvers with an accurate ranking.
We define the \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) by the fraction of pairs $\left(\hat{a}, a\right)$ for all $a \in \mathcal{A}$ that are decided correctly by the given ranking (cf.~Algorithm~\ref{alg:eval}).
For now, we exclude the equality of solvers since data shows that the PAR-2 scores of all our solvers are different.
So, possible ranking decisions may be, for example, solver $a$ is better than $b$ or $b$ is better than $a$.
The ranking accuracy is affected by whether this decision is correctly made.
%Matricon et. al.~\cite{MatriconAFSH21} also evaluate their approach by the fraction of solver pairs that is correctly decided.
Second, we also have to optimize for runtime.
The \emph{fraction of runtime} that the algorithm needs to arrive at its conclusion is denoted by $O_{\operatorname{rt}} \in \left[0, 1\right]$ (lower is better).
This metric puts the runtime summed over the sampled instances in relation to the runtime summed over all instances in the dataset (cf. Algorithm~\ref{alg:eval}).
Overall, we want to find an approach that maximizes
%
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
%
whereby $\delta \in \left[0, 1\right]$ allows for linear weighting between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
Plotting the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on a $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.

\subsection{Evaluation Framework}
\label{sec:exdesign:evalframe}

@@

\subsection{Hyper-parameters}
\label{sec:exdesign:hyper}

Given Algorithm~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}.
Also, there are different choices for the runtime-label prediction model.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results (cf. Section~\ref{sec:eval:prediction}) and do the main end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.

@@

Regarding \emph{ranking} (cf. Section~\ref{sec:main:model}), we experiment with the following approaches, including our used hyper-parameter values:

\begin{itemize}
  \item Observed PAR-2 ranking of already sampled instances
  \item Predicted ranking induced by runtime-label predictions
  \begin{itemize}
    \item
    History size: consider the latest 1, 10, 20, 30, or 40 predictions within a voting approach for stability.
    The latest $x$ predictions vote on the winner.
    %JB: confuses me a bit as it sounds like we rank first and then vote on a winner; isn't it the other way around?
    %TF: Removed confusing sentence
    \item
    Fallback threshold: if the difference of scores between two solvers drops below \SI{0.01}, \SI{0.05}, or \SI{0.1}, use the partially observed PAR-2 ranking as a tie-breaker.
    %JB: as clarification: one these two solvers has to be the new solver, right? if yes, would write it explicitly
    %TF: Not necessarily
    As discussed in Section~\ref{sec:main:model}, ranking decisions are all correct if the difference in scores between two solvers exceeds a certain value.
  \end{itemize}
\end{itemize}

@@

For \emph{selection} (cf. Section~\ref{sec:main:sampling}), we experiment with the following methods, including our used hyper-parameter values:

@@

For \emph{stopping} decisions (cf. Section~\ref{sec:main:stopping}), we experiment with the following criteria, including our used hyper-parameter values:

@@

As runtime-label prediction model, we only make use of one fixed model since an exhaustive grid search would be infeasible (cf.~Section~\ref{sec:eval:prediction}).
Our model of choice is an ensemble, stacking a quadratic-discriminant analysis\footnote{With a regularization parameter of zero and a singular value threshold of zero} onto a random forest model\footnote{With \emph{entropy} splitting criterion and \emph{balanced} class weights} using a decision tree\footnote{With \emph{Gini} impurity splitting criterion; choosing the best split for each branching decision and maximum tree depth of 5 levels} to weight its members.
If not stated explicitly, all other model parameters default to the presets of \emph{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/index.html}; Version 1.0.2}~\cite{scikit-learn}.

@@

\subsection{Runtime Prediction}
\label{sec:eval:prediction}

@@

For runtime discretization, we use hierarchical clustering with $k = 3$ and a log-single-link criterion for all further experiments.
Each non-time-out runtime starts in a separate interval.
We then gradually merge intervals whose single-link logarithmic distance is the smallest until the desired number of partitions is reached.
As already mentioned in Section~\ref{sec:main:model}, this approach has good properties regarding the predictiveness of the PAR-2 score ranking and its discriminatory power.
Also, experiments showed that our stacking ensemble performed particularly well for this kind of discretization method.
Other clustering approaches that we have tried include hierarchical clustering with mean-, median- and complete-link criteria, as well as $k$-means and spectral clustering.
%Predicting runtime labels in a similar experiment as above produced an MCC score of on average \SI{88.82}{\%} ($\sigma = \SI{4.75}{\%}$; in comparison to \SI{95.27}{\%} for predicting time-outs).
%JB: "similar experiment as above" - reference unclear; what do you want to say?
%TF: Removed the sentence

@@

Our end-to-end experiments follow the evaluation framework introduced in Section~\ref{sec:exdesign:evalframe}.
Fig.~\ref{fig:e2eallsolvers} shows the performance of the approaches from Section~\ref{sec:exdesign:hyper} on $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams for the smaller hyper-parameter-tuning dataset.
Evaluating a particular configuration with Algorithm~\ref{alg:eval} returns a point $\left(O_{\operatorname{rt}},\, O_{\operatorname{acc}}\right)$.
The plotted lines represent the best performing configurations (convex hull) per top-level hyper-parameter choice, i.e., per ranking approach (Fig.~\ref{fig:annitraincolorranking}), selection approach (Fig.~\ref{fig:annitraincolorselection}), and stopping criterion (Fig.~\ref{fig:annitraincolorstopping}).
%Colors indicate the different hyper-parameter instantiations.
%To be more specific, they represent the Pareto front of all configurations with the respective instantiations.

@@

  \caption{
  	$O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagrams comparing different hyper-parameter instantiations of our active-learning framework.
  	The x-axis shows the ratio of total solver runtime on the sampled instances relative to all instances.
  	The y-axis shows the ranking accuracy (cf. Section~\ref{sec:exdesign:goal}).
  	%JB: we should use "ranking accuracy" in plots, since we explicitly defined this metric earlier
  	Each line entails the front of Pareto-optimal configurations for the respective hyper-parameter instantiation.
  }
  \label{fig:e2eallsolvers}
\end{figure}
